Human Factors Influencing Visual Statistical Inference

### 2014-10-24 sociology_paper.tex
factors - halfway

Because visual statistical inference relies on plotting choices, lineup design and human evaluation, it is important to understand how these factors that might affect results.  This section provides a brief description of the factors that are expected to have some impact on the performance of visual statistical inference.

@@

\begin{itemize} \itemsep 0in
%\subsection{Choice of Visual Test Statistic}  
\item{\bf Choice of Visual Test Statistic:}  
There are a variety of ways that data can be plotted and for plots to be enhanced by axes, legends or representations of models. The choice of plot chosen as the visual test statistic, primarily should be appropriate for the problem being investigated. For example, for studying the association between two quantitative variables a scatterplot is usually optimal. Enhancements, such as a line overlaying a scatterplot to represent a fitted model, might be appropriate for studying models for the data.  Some choices of visual test statistics provide better results than others when studying the same problem, which is discussed in \cite{heike:2012}.  This makes sense, because some plot types will make it easier for certain structures to be seen.

@@

%\subsection{Signal in the Data}
\item{\bf Signal in the Data:}
A visual test statistic is chosen for a particular task. If it is well-designed then departures from the null hypothesis will be visible in the plot, especially if the signal is strong. An important factor that helps an observer to identify the actual data plot in a lineup is the strength of the signal in the data. Simulated data was generated for all of the experiments, except 9, enabling the study of signal strength and actual data plot detection. Primarily the results were as expected, that as the signal strength increased observers more frequently picked the actual data plot as different from the null plots.

@@

%\subsection{Question Design} 
\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of structure, enabling the the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the observer is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations. (\citet{buja:2009} has a very complete explanation.)

In the initial Turk experiments (1,2,3) conducted with simulated data to compare the results from visual inference with those obtained by classical inference, the questions posed to the observers were more targeted. This was important in order to make a comparison of the results. For latter studies (4, 7, 10), question posed to the observer was broader. Table \ref{tbl:visual_stat} lists the questions used for each experiment.

@@

%\subsection{Demographics} 
\item{\bf Demographics:}  The experiments collected data on age, gender, education level and geographic location. Each observer self-reported their gender, age in roughly five year intervals, education level as high school or less, some college courses, an undergraduate degree, some graduate courses or a graduate degree. The IP address of the computer afforded the geographical location of the subject. The purpose of collecting this information with each experiment~is to examine the effect that they have on the results of visual inference -- ideally very little.

>>>>

\hh{Visual statistical inference relies not only on the strength of the signal in the data, but also on how this information is presented in the plot, on the lineup design and on the abilities of human observers.} It is important to understand how these factors might affect results.  \hh{Here, we provide a brief discussion }%This section provides a brief description 
of the factors that are expected to have some impact on the performance of visual statistical inference.

@@

\begin{itemize} \itemsep 0in \parsep-10pt %\parskip0pt 
%\subsection{Choice of Visual Test Statistic}  
\item{\bf Choice of Visual Test Statistic:}  
%There are a variety of ways that
Data can be plotted in a variety of ways and we can enhance plots by axes, legends or representations of models. The choice of plot primarily should be appropriate for the problem being investigated, as the data plot is the visual test statistic. 
When for example studying the association between two quantitative variables a scatterplot is usually optimal. \hh{XXX do we really want to say optimal? that sounds like we have an underlying criterion to optimize. XXX} Enhancements, such as a line overlaying a scatterplot to represent a fitted model, might be appropriate for studying models for the data.  Some choices of visual test statistics provide better results than others when studying the same problem \citep{heike:2012}, because some plot types make it easier for certain structures to be seen.

@@

%\subsection{Signal in the Data}
\item{\bf Signal in the Data:}
A visual test statistic is chosen for a particular task. If it is well-designed then departures from the null hypothesis will be visible in the plot, especially if the signal is strong. An important factor that helps an observer to identify the actual data plot in a lineup is the strength of the signal in the data. Simulated data was generated for all of the experiments, except exps.~4 and~9, enabling the study of signal strength and actual data plot detection. \hh{In experiment~4 we took a different approach by varying the sampling size, which also allowed us to investigate the effect a design had on detection rate while varying signal/noise ratio.} Primarily the results were as expected, that as the signal strength increased observers more frequently picked the actual data plot as different from the null plots.

@@

%\subsection{Question Design} 
\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of structure, enabling the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the observer is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations \citet[see also ][]{buja:2009}.

In the initial Turk experiments (1,2,3) conducted with simulated data to compare the results from visual inference with those obtained by classical inference, the questions posed to the observers were more targeted. This was important \hh{to allow a direct comparison of the results to results from conventional testing.}%in order to make a comparison of the results. 
For latter studies (4, 7, 10), questions posed to the observer were more general. Table \ref{tbl:visual_stat} lists the questions used for each experiment.

@@

%\subsection{Demographics} 
\item{\bf Demographics:}  The experiments collected data on age, gender, education level and geographic location. Each observer self-reported gender, age in roughly five year intervals, education level as high school or less, some college courses, an undergraduate degree, some graduate courses or a graduate degree. The IP address of the computer afforded the geographical location of the subject. The purpose of collecting this information with each experiment is to examine the effect that they have on the results of visual inference -- ideally very little.
